Logical Cryptanalysis Benchmarks for \\ Classical and Modern Hash Functions

### 2018-04-16 sat-competition-2018/conference.tex
Spellcheck
Signed-off-by: Alexander Scheel <alexander.m.scheel@gmail.com>

\section{Introduction}
This collection of benchmarks focuses on new techniques in logical
cryptanalysis for analyzing the structure of various hash functions. We
offer as a benchmark a new technique analyzing the security of a hash function
of the Merkle-D{\aa}mgard construction which are broadly applicable to all such
constructions and several benchmarks analyzing the various security properties
of the Keccak hash function. While the latter techniques have not yeilded
useful cryptanalysis results, they pose as a source of variable scalable
benchmarks due to several choices of parameters (the bitwidth, $w$ and the
number of rounds).

@@

The notion of a differential family is not as useful for Keccak due to
the interaction between the sponge function and the internal permutation
functions causing the collision space of families to be entirely different
between successive rounds. From an algebraic perspective, and to study the
security margin of the XOF consruct \cite{Keccak3}, it is important to show
that the Keccak round functions are bijective and have high permutation order.
Two of our benchmarks (\texttt{bijection} and \texttt{orders}) model these
problems in SAT. In general, these instances are easy, but with a few notable
outliers: for $w \geq 4$, proving the order of $\theta$ is $3w$ is difficult.
Note that this should simplify to showing that $\forall x, x \neq x$ is UNSAT,
and thus should be relatively trivial; however $w=4$ produces runtimes in
excess of one hour. Further, anything later involving $\theta$ (such as
$\rho \circ \theta$, etc.) also becomes difficult for $w\geq4$.

In \texttt{differences}, we study the differential properties of the
permutation functions: for a given parameters $x, y \leq 25w$ and round
function $f$, we seek to find witnesses $a$ and $b$ such that:
\begin{align*}
    & \#(1, a \oplus b) = x \\
    & \#(1, f(a) \oplus f(b)) = y
\end{align*}
That is, find an input with difference $x$ which produces an output with
difference $y$. There are a few interesting outliers in this model: for
$w\geq4$, most round functions cannot be individually analyzed this way.
Further, while $\pi$ is a permutation of the order of the bits (and thus
does not change the values of any bits), certain instances in $w=2$
where $x \neq y$ produce runtimes in excess of an hour. This is surprising
as the model is trivially SAT if and only if $x == y$ (and the model merely
involves changing the orders of variables).

The benchmarks in \texttt{output-margins} consider the effects of the sponge
function. Since all of the round functions are bijective, if two inputs differ,
the interal state of Keccak must also differ. However, to produce a collision,
only the first $y$ bits of the output must be the same. Thus we can create
a model for a given number of differences $x, z \leq 25w$, security margin
$y \leq 25w$, and round functions $f$:
\begin{align*}
    & \#(1, a xor b) = x \\
    & \#(1, (f(a) xor f(b))[0:y]) = z
\end{align*}
If such a witness a and b exist for $z=0$, then the input difference $x$ is
possible of producing collisions at a security margin $m$. Our provided
benchmarks sample the space for small values of $w$ and relatively few
round functions; for $w\geq 16$ and for any set of functions including
$\theta$, the runtimes become exceedingly long.

The benchmarks in \texttt{xof-state} attempt to recreate the internal state
of Keccak given a series of outputs from the XOF (extensible output function
at a given margin). In general, this is possible for either small values of
$w$ or small numbers of rounds (for larger values of $w$). However, care must
be selected in choosing the base seed, otherwise, there can possibly be
multiple satisfying seeds. However, for a set of inputs with unique solution,
these benchmarks can be extended to contain the output from several rounds of
Keccak. After a threshhold dependent on the margin, these are redundant
information and thus test the SAT solvers to work with larger models which
overfit to the solution.

>>>>

\section{Introduction}
This collection of benchmarks focuses on new techniques in logical
cryptanalysis for analyzing the structure of various hash functions. We
offer as a benchmark a new technique analyzing the security of a hash function
of the Merkle-D{\aa}mgard construction which are broadly applicable to all such
constructions and several benchmarks analyzing the various security properties
of the Keccak hash function. While the latter techniques have not yielded
useful cryptanalysis results, they pose as a source of variable scalable
benchmarks due to several choices of parameters (the bit width, $w$ and the
number of rounds).

@@

The notion of a differential family is not as useful for Keccak due to
the interaction between the sponge function and the internal permutation
functions causing the collision space of families to be entirely different
between successive rounds. From an algebraic perspective, and to study the
security margin of the XOF construct \cite{Keccak3}, it is important to show
that the Keccak round functions are bijective and have high permutation order.
Two of our benchmarks (\texttt{bijection} and \texttt{orders}) model these
problems in SAT. In general, these instances are easy, but with a few notable
outliers: for $w \geq 4$, proving the order of $\theta$ is $3w$ is difficult.
Note that this should simplify to showing that $\forall x, x \neq x$ is UNSAT,
and thus should be relatively trivial; however $w=4$ produces run times in
excess of one hour. Further, anything later involving $\theta$ (such as
$\rho \circ \theta$, etc.) also becomes difficult for $w\geq4$.

In \texttt{differences}, we study the differential properties of the
permutation functions: for a given parameters $x, y \leq 25w$ and round
function $f$, we seek to find witnesses $a$ and $b$ such that:
\begin{align*}
    & \#(1, a \oplus b) = x \\
    & \#(1, f(a) \oplus f(b)) = y
\end{align*}
That is, find an input with difference $x$ which produces an output with
difference $y$. There are a few interesting outliers in this model: for
$w\geq4$, most round functions cannot be individually analyzed this way.
Further, while $\pi$ is a permutation of the order of the bits (and thus
does not change the values of any bits), certain instances in $w=2$
where $x \neq y$ produce run times in excess of an hour. This is surprising
as the model is trivially SAT if and only if $x == y$ (and the model merely
involves changing the orders of variables).

The benchmarks in \texttt{output-margins} consider the effects of the sponge
function. Since all of the round functions are bijective, if two inputs differ,
the internal state of Keccak must also differ. However, to produce a collision,
only the first $y$ bits of the output must be the same. Thus we can create
a model for a given number of differences $x, z \leq 25w$, security margin
$y \leq 25w$, and round functions $f$:
\begin{align*}
    & \#(1, a xor b) = x \\
    & \#(1, (f(a) xor f(b))[0:y]) = z
\end{align*}
If such a witness a and b exist for $z=0$, then the input difference $x$ is
possible of producing collisions at a security margin $m$. Our provided
benchmarks sample the space for small values of $w$ and relatively few
round functions; for $w\geq 16$ and for any set of functions including
$\theta$, the runtimes become exceedingly long.

The benchmarks in \texttt{xof-state} attempt to recreate the internal state
of Keccak given a series of outputs from the XOF (extensible output function
at a given margin). In general, this is possible for either small values of
$w$ or small numbers of rounds (for larger values of $w$). However, care must
be selected in choosing the base seed, otherwise, there can possibly be
multiple satisfying seeds. However, for a set of inputs with unique solution,
these benchmarks can be extended to contain the output from several rounds of
Keccak. After a threshold dependent on the margin, these are redundant
information and thus test the SAT solvers to work with larger models which
over fit to the solution.
